---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mwi

<!-- badges: start -->
[![R-CMD-check](https://github.com/E4Warning/mwi/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/E4Warning/mwi/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The Mosquito Weather Index (MWI) combines air temperature, relative humidity
and wind speed into a single number between 0 and 1 describing how favourable
the weather is for adult mosquito activity, and maps that number onto five
ordered activity classes. It was developed by the Benaki Phytopathological
Institute and the National Observatory of Athens under the LIFE CONOPS project
(LIFE12 ENV/GR/000466) and has been published operationally on
[meteo.gr](https://meteo.gr) since June 2014.

This package implements the index, the three response functions it is built
from, its activity classes, and the unit conversions needed to feed it from
common weather products. It has no dependencies beyond base R.

Full documentation, including a worked introduction, is at
<https://e4warning.github.io/mwi/>.

## Installation

```r
# install.packages("pak")
pak::pak("E4Warning/mwi")
```

## Usage

```{r}
library(mwi)

mwi(temperature = 22, humidity = 70, wind_speed = 10)
```

The index is a product of three responses, so any one variable falling outside
its favourable range sets the whole index to zero:

```{r}
mwi(temperature = 22, humidity = 30, wind_speed = 10) # too dry
mwi(temperature = 35, humidity = 70, wind_speed = 10) # too hot
mwi(temperature = 22, humidity = 70, wind_speed = 30) # too windy
```

The individual responses are available on their own:

```{r}
mwi_temperature(c(10, 17.5, 22, 27.5, 32))
mwi_humidity(c(30, 40, 67.5, 95))
mwi_wind(c(10, 21.6, 25))
```

Continuous values map onto the five classes used for public communication:

```{r}
x <- c(0, 0.2, 0.5, 0.9, 1)
data.frame(mwi = x, class = mwi_category(x))
```

## Computing the index over time

**The index must be computed before it is averaged, not after.** Because `mwi()`
is non-linear, evaluating it on daily summaries of each weather variable is not
the same as evaluating it hourly and then summarising the result — and it is
much less useful. Daily maxima of temperature and of humidity typically occur
many hours apart, so an index built from them describes an hour that never
happened.

`mwi_summarise()` does it in the right order:

```{r}
hours <- 0:23
temp <- 18 + 8 * sin((hours - 6) / 24 * 2 * pi)
rh <- 75 - 15 * sin((hours - 6) / 24 * 2 * pi)
wind <- rep(5, 24)

# Right: index each hour, then take the daily maximum
mwi_summarise(temp, rh, wind, by = rep("2019-07-15", 24))

# Wrong: summarise the weather first, then index
mwi(max(temp), max(rh), max(wind))
```

## Working from reanalysis output

Products such as ERA5-Land distribute temperatures in Kelvin, humidity as a
dewpoint, and wind as orthogonal components. The conversions are included:

```{r}
temperature <- celsius_from_kelvin(295.15)
dewpoint <- celsius_from_kelvin(288.15)
humidity <- relative_humidity_from_dewpoint(dewpoint, temperature)
wind <- wind_speed_from_components(u = 1.2, v = -0.8)

mwi(temperature, humidity, wind, units = "m/s")
```

## Citation

If you use this package, please cite the evaluation study alongside it. See
`citation("mwi")`.

## License

GPL (>= 3). The index itself was developed by the Benaki Phytopathological
Institute and the National Observatory of Athens under the LIFE CONOPS project
(LIFE12 ENV/GR/000466).
