---
title: "Getting started with the Mosquito Weather Index"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with the Mosquito Weather Index}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(mwi)
```

## What the index is

The Mosquito Weather Index reduces three weather variables to one number in
`[0, 1]`. It is the product of three response functions, each of which returns
a suitability multiplier for one variable:

```{r}
mwi_temperature(22)
mwi_humidity(70)
mwi_wind(10)

mwi_temperature(22) * mwi_humidity(70) * mwi_wind(10)
mwi(temperature = 22, humidity = 70, wind_speed = 10)
```

The shapes are worth knowing, because they explain most of the index's
behaviour.

`mwi_temperature()` is a trapezoid, flat and maximal from 20 to 25 °C and zero
outside 15–30 °C:

```{r}
t <- seq(10, 35, by = 5)
data.frame(temperature = t, response = mwi_temperature(t))
```

`mwi_humidity()` ramps linearly from 40% to 95%, then drops to zero as a step:

```{r}
h <- c(35, 40, 60, 80, 95, 96)
data.frame(humidity = h, response = mwi_humidity(h))
```

That step at 95% is deliberate, not an artefact. It has a practical
consequence: the index reaches 1 only when humidity is *exactly* 95%, so the
top "very high activity" class is attained on rounded observational data but
almost never on continuous data such as reanalysis output.

`mwi_wind()` is a switch rather than a gradient:

```{r}
mwi_wind_threshold()
mwi_wind(c(21.5, 21.6, 21.7))
```

Because the three responses are multiplied, any one of them hitting zero takes
the whole index with it. Wind and humidity are therefore best thought of as
gates, and temperature as the thing that grades the result.

## Compute first, aggregate second

This is the part that most often goes wrong.

`mwi()` is non-linear, so it does not commute with averaging. If you have
hourly weather and you want a daily index, you must evaluate the index hourly
and then summarise the resulting index values. Summarising the weather first
and then indexing gives a different answer.

The problem is worst with daily maxima, because the daily maximum of relative
humidity and the daily maximum of temperature fall at opposite ends of the day —
humidity peaks in the small hours, temperature in mid-afternoon. Combining them
describes an hour that never occurred.

```{r}
hours <- 0:23
temperature <- 18 + 8 * sin((hours - 6) / 24 * 2 * pi)
humidity <- 75 - 15 * sin((hours - 6) / 24 * 2 * pi)
wind_speed <- rep(5, 24)

# Note the phase difference: they peak twelve hours apart
c(warmest_hour = which.max(temperature) - 1,
  most_humid_hour = which.max(humidity) - 1)
```

```{r}
# Right: index each hour, then reduce
mwi_summarise(temperature, humidity, wind_speed,
              by = rep("2019-07-15", 24), statistic = "max")

# Wrong: reduce the weather, then index
mwi(max(temperature), max(humidity), max(wind_speed))
```

`mwi_summarise()` enforces the correct order. Group with `by` to get one row
per day:

```{r}
two_days <- rep(as.Date(c("2019-07-15", "2019-07-16")), each = 24)
mwi_summarise(rep(temperature, 2), rep(humidity, 2), rep(wind_speed, 2),
              by = two_days)
```

### Maximum or mean?

Both are available through `statistic`. In the evaluation study accompanying
this package, the daily *maximum* consistently predicted adult trap counts
better than the daily mean, which is consistent with catch responding to
whether the day contained a window of favourable conditions rather than to how
favourable the day was on average.

```{r}
by <- rep("2019-07-15", 24)
rbind(
  mwi_summarise(temperature, humidity, wind_speed, by, statistic = "max"),
  mwi_summarise(temperature, humidity, wind_speed, by, statistic = "mean"),
  mwi_summarise(temperature, humidity, wind_speed, by, statistic = "min")
)
```

## Activity classes

For communication the continuous index is reduced to five ordered classes:

```{r}
x <- c(0, 0.15, 0.5, 0.85, 1)
data.frame(mwi = x, class = mwi_category(x))
```

The result is an ordered factor, so it can be compared and sorted:

```{r}
mwi_category(0.85) > mwi_category(0.15)
```

Pass `labels` to translate them:

```{r}
greek <- c("Καμία", "Χαμηλή", "Μέτρια", "Υψηλή", "Πολύ υψηλή")
mwi_category(c(0, 0.5, 1), labels = greek)
```

## Feeding the index from reanalysis data

Gridded products rarely give you the three variables in the form the index
wants. ERA5-Land, for instance, distributes 2 m air temperature and 2 m
dewpoint in Kelvin, and 10 m wind as eastward and northward components in m/s.
The conversions are included:

```{r}
t2m <- c(295.15, 299.15, 291.15)
d2m <- c(288.15, 289.15, 287.15)
u10 <- c(1.2, -3.4, 0.5)
v10 <- c(-0.8, 2.1, 1.1)

temperature <- celsius_from_kelvin(t2m)
dewpoint <- celsius_from_kelvin(d2m)
humidity <- relative_humidity_from_dewpoint(dewpoint, temperature)
wind_speed <- wind_speed_from_components(u10, v10)

data.frame(
  temperature = round(temperature, 1),
  humidity = round(humidity, 1),
  wind_ms = round(wind_speed, 1),
  mwi = round(mwi(temperature, humidity, wind_speed, units = "m/s"), 3)
)
```

## When wind is unavailable

Temperature and humidity loggers do not measure wind. `mwi_fhft()` drops the
wind term, which is equivalent to assuming wind is always below the cut-off:

```{r}
mwi_fhft(temperature = 22, humidity = 70)
mwi(temperature = 22, humidity = 70, wind_speed = 0)
```

Treat this variant with care. The evaluation study found that substituting
on-site logger readings for gridded reanalysis made the index a *worse*
predictor of trap counts, partly because logger humidity ran systematically dry
and the 40% threshold amplifies that bias.

## Missing values

Every function propagates `NA` rather than guessing:

```{r}
mwi(temperature = c(22, NA), humidity = 70, wind_speed = 5)
```

In `mwi_summarise()`, a group containing any missing index value returns `NA`
unless you ask otherwise:

```{r}
gappy_temperature <- 18 + 8 * sin((hours - 6) / 24 * 2 * pi)
gappy_temperature[5] <- NA
gappy_humidity <- 75 - 15 * sin((hours - 6) / 24 * 2 * pi)

mwi_summarise(gappy_temperature, gappy_humidity, rep(5, 24), by)$mwi
mwi_summarise(gappy_temperature, gappy_humidity, rep(5, 24), by,
              na.rm = TRUE)$mwi
```

Note that arguments must be the same length or length 1. Partial recycling is
an error rather than a silent surprise:

```{r, error = TRUE}
mwi(temperature = c(20, 22, 24, 26), humidity = c(60, 70), wind_speed = 5)
```
